\bf{Supplementary Material for ``Longitudinal Principal Manifold Estimation"}

### 2025-02-10 manuscript/paper/main.tex
reset to JASA submission version, add title/abstract revisions

\documentclass[12pt]{article}
\usepackage{algorithm2e}
\usepackage{amscd}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{authblk}
\usepackage[english]{babel}
\usepackage{bm}
\usepackage{caption}
\usepackage{color}
\usepackage{float}
\usepackage[perpage, symbol*]{footmisc}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{enumerate}
\usepackage{mathrsfs}
\usepackage{natbib}
\usepackage{setspace}
\usepackage{subcaption}
\usepackage{tikz}
\usepackage{url}

@@

@@

\if1\blind
{
  \title{\bf Longitudinal Principal Manifold Estimation}
  \author{
    Robert Zielinski\hspace{.2cm}\\
    Department of Biostatistics, Brown University\\
    Kun Meng \\
    Division of Applied Mathematics, Brown University\\
    and \\
    Ani Eloyan \\
    Department of Biostatistics, Brown University\\
    for the Alzheimer's Disease Neuroimaging Initiative\thanks{Data used in preparation of this article were obtained from the Alzheimer's Disease Neuroimaging Initiative (ADNI) database (adni.loni.usc.edu). As such, the investigators within the ADNI contributed to the design and implementation of ADNI and/or provided data but did not participate in analysis or writing of this report. A complete listing of ADNI investigators can be found at \href{http://adni.loni.usc.edu/wp-content/uploads/how_to_apply/ADNI_Acknowledgement_List.pdf}{\texttt{http://adni.loni.usc.edu/wp-content/uploads/how\_to\_apply/ADNI\_Acknowledgement\_List.pdf}}}}
  \maketitle
} \fi

@@

\bigskip
\begin{abstract}
Longitudinal magnetic resonance imaging data is used to model trajectories of change in brain regions of interest to identify areas susceptible to atrophy in those with neurodegenerative conditions like Alzheimer’s disease. Most methods for extracting brain regions are applied to scans from study participants independently, resulting in wide variability in shape and volume estimates of these regions over time in longitudinal studies. To address this problem, we propose a longitudinal principal manifold estimation method, which seeks to recover smooth, longitudinally meaningful manifold estimates of shapes over time. The proposed approach uses a smoothing spline to smooth over the coefficients of principal manifold embedding functions estimated at each time point. This mitigates the effects of random disturbances to the manifold between time points. Additionally, we propose a novel data augmentation approach to enable principal manifold estimation on self-intersecting manifolds. Simulation studies demonstrate performance improvements over naïve applications of principal manifold estimation and principal curve/surface methods. The proposed method improves the estimation of surfaces of hippocampuses and thalamuses using data from participants of the Alzheimer’s Disease Neuroimaging Initiative. An analysis of magnetic resonance imaging data from 236 individuals shows the advantages of our proposed methods that leverage regional longitudinal trends for segmentation.
\end{abstract}

>>>>

\documentclass[12pt]{article}
\usepackage{algorithm2e}
\usepackage{amscd}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{authblk}
\usepackage[english]{babel}
\usepackage{bm}
\usepackage{caption}
\usepackage{color}
\usepackage{float}
\usepackage[perpage, symbol*]{footmisc}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{enumerate}
\usepackage{mathrsfs}
\usepackage{natbib}
\usepackage{setspace}
\usepackage{subcaption}
\usepackage{tikz}
\usepackage{ulem}
\usepackage{url}

@@

\newcommand{\revision}[1]{\textcolor{blue}{#1}}
\newcommand{\outdated}[1]{\sout{#1}}

@@

\if1\blind
{
  \title{\bf \outdated{Longitudinal Principal Manifold Estimation} \revision{Noise-corrected hippocampal shape and volume trajectories in Alzheimer's disease using longitudinal principal manifold estimation}}
  \author{
    Robert Zielinski\hspace{.2cm}\\
    Department of Biostatistics, Brown University\\
    Kun Meng \\
    Division of Applied Mathematics, Brown University\\
    and \\
    Ani Eloyan \\
    Department of Biostatistics, Brown University\\
    for the Alzheimer's Disease Neuroimaging Initiative\thanks{Data used in preparation of this article were obtained from the Alzheimer's Disease Neuroimaging Initiative (ADNI) database (adni.loni.usc.edu). As such, the investigators within the ADNI contributed to the design and implementation of ADNI and/or provided data but did not participate in analysis or writing of this report. A complete listing of ADNI investigators can be found at \href{http://adni.loni.usc.edu/wp-content/uploads/how_to_apply/ADNI_Acknowledgement_List.pdf}{\texttt{http://adni.loni.usc.edu/wp-content/uploads/how\_to\_apply/ADNI\_Acknowledgement\_List.pdf}}}}
  \maketitle
} \fi

@@

\bigskip
\begin{abstract}
Longitudinal \revision{structural} magnetic resonance imaging data is used to model trajectories of change in brain regions of interest to identify areas susceptible to atrophy in those with neurodegenerative conditions like Alzheimer’s disease. Most methods for extracting brain regions are applied to scans from study participants independently, resulting in wide variability in shape and volume estimates of these regions over time in longitudinal studies. To address this problem, we propose a longitudinal principal manifold estimation method, which seeks to recover smooth, longitudinally meaningful manifold estimates of shapes over time. The proposed approach uses a smoothing spline to smooth over the coefficients of principal manifold embedding functions estimated at each time point. This mitigates the effects of random disturbances to the manifold between time points. Additionally, we propose a novel data augmentation approach to enable principal manifold estimation on self-intersecting manifolds. Simulation studies demonstrate performance improvements over naïve applications of principal manifold estimation and principal curve/surface methods. The proposed method improves the estimation of surfaces of \outdated{hippocampuses} \revision{hippocampi} and \outdated{thalamuses} \revision{thalami} using data from participants of the Alzheimer’s Disease Neuroimaging Initiative. An analysis of magnetic resonance imaging data from 236 individuals shows the advantages of our proposed methods that leverage regional longitudinal trends for segmentation.
\end{abstract}

### 2025-02-10 manuscript/paper/supplement.tex
reset to JASA submission version, add title/abstract revisions

  \begin{subfigure}{\textwidth}
    \label{fig:adni_lhipp_result}
    \begin{subfigure}{\textwidth}
      \centering
      \includegraphics[height=3cm]{adni_plots/adni_lhipp_data_plot}
      \caption{Data}
    \end{subfigure}
    \begin{subfigure}{\textwidth}
      \centering
      \includegraphics[height=3cm]{adni_plots/adni_lhipp_lpme_isomap_plot}
      \caption{LPME}
    \end{subfigure}
    \begin{subfigure}{\textwidth}
      \centering
      \includegraphics[height=3cm]{adni_plots/adni_lhipp_pme_plot}
      \caption{PME}
    \end{subfigure}   
  \end{subfigure}

  \begin{subfigure}{\textwidth}
    \label{fig:adni_lthal_result}
    \begin{subfigure}{\textwidth}
      \centering
      \includegraphics[height=3cm]{adni_plots/adni_lthal_data_plot}
      \caption{Data}
    \end{subfigure}
    \begin{subfigure}{\textwidth}
      \centering
      \includegraphics[height=3cm]{adni_plots/adni_lthal_lpme_isomap_plot}
      \caption{LPME}
    \end{subfigure}
    \begin{subfigure}{\textwidth}
      \centering
      \includegraphics[height=3cm]{adni_plots/adni_lthal_pme_plot}
      \caption{PME}
    \end{subfigure}   
  \end{subfigure}

@@

\begin{figure}
    \centering
    \subfloat[\centering Left Hippocampus Volume Estimates]{
      \includegraphics[height=8cm]{adni_plots/adni_lhipp_volume_comp}
      \label{fig:lhipp_volume_comparison}
    }
    \vfill
    \subfloat[\centering Left Thalamus Volume Estimates]{
      \includegraphics[height=8cm]{adni_plots/adni_lthal_volume_comp}
      \label{fig:lthal_volume_comparison}
    }
    \caption{Left hippocampus and left thalamus volume estimates for three ADNI participants. Volume estimates are obtained by summing the volumes of voxels contained within the structure as described by the segmented data (circles, solid line), LPME (squares, dashed line), and PME (triangles, dotted line). When PME and LPME are applied to the irregular shape of the hippocampus, there is a clear ordering of the volume estimates. Gaps in the estimated surface induce underestimates of the volume compared to the volume values estimated from the data. When applied to the thalamus, the PME-based volume estimates demonstrate similar time point-to-time point changes in the volumes estimated from the segmented data, reflecting the close fit of the PME algorithm to the regularly-shaped thalamus data. The LPME estimates appear to successfully smooth over regions with large variations in subsequent volume measures from the data and PME estimates, as seen for participant 033\_S\_0514.}
\end{figure}

>>>>

  \begin{subfigure}{\textwidth}
    \label{fig:adni_lhipp_result}
    \begin{subfigure}{\textwidth}
      \centering
      \includegraphics[height=3cm]{adni_lhipp_data_plot}
      \caption{Data}
    \end{subfigure}
    \begin{subfigure}{\textwidth}
      \centering
      \includegraphics[height=3cm]{adni_lhipp_lpme_isomap_plot}
      \caption{LPME}
    \end{subfigure}
    \begin{subfigure}{\textwidth}
      \centering
      \includegraphics[height=3cm]{adni_lhipp_pme_plot}
      \caption{PME}
    \end{subfigure}
  \end{subfigure}

  \begin{subfigure}{\textwidth}
    \label{fig:adni_lthal_result}
    \begin{subfigure}{\textwidth}
      \centering
      \includegraphics[height=3cm]{adni_lthal_data_plot}
      \caption{Data}
    \end{subfigure}
    \begin{subfigure}{\textwidth}
      \centering
      \includegraphics[height=3cm]{adni_lthal_lpme_isomap_plot}
      \caption{LPME}
    \end{subfigure}
    \begin{subfigure}{\textwidth}
      \centering
      \includegraphics[height=3cm]{adni_lthal_pme_plot}
      \caption{PME}
    \end{subfigure}
  \end{subfigure}

@@

\begin{figure}
    \centering
    \subfloat[\centering Left Hippocampus Volume Estimates]{
      \includegraphics[height=8cm]{adni_lhipp_volume_comp}
      \label{fig:lhipp_volume_comparison}
    }
    \vfill
    \subfloat[\centering Left Thalamus Volume Estimates]{
      \includegraphics[height=8cm]{adni_lthal_volume_comp}
      \label{fig:lthal_volume_comparison}
    }
    \caption{Left hippocampus and left thalamus volume estimates for three ADNI participants. Volume estimates are obtained by summing the volumes of voxels contained within the structure as described by the segmented data (circles, solid line), LPME (squares, dashed line), and PME (triangles, dotted line). When PME and LPME are applied to the irregular shape of the hippocampus, there is a clear ordering of the volume estimates. Gaps in the estimated surface induce underestimates of the volume compared to the volume values estimated from the data. When applied to the thalamus, the PME-based volume estimates demonstrate similar time point-to-time point changes in the volumes estimated from the segmented data, reflecting the close fit of the PME algorithm to the regularly-shaped thalamus data. The LPME estimates appear to successfully smooth over regions with large variations in subsequent volume measures from the data and PME estimates, as seen for participant 033\_S\_0514.}
\end{figure}
